---
title: "FINALPROJECT"
author: "Andrea Perez Pizarro"
editor: visual
bibliography: references.bib
csl: apa.csl  
---

```{r initial, message=FALSE, warning=FALSE, echo=FALSE}
library(dplyr)
library(readr)

# Load data
cleandata <- read_csv("cleandata.csv")
hogar_data <- read_csv("hogar_ensanut.csv")

# Ensure FOLIO_I is of the same type
cleandata <- cleandata %>% mutate(FOLIO_I = as.character(FOLIO_I))
hogar_data <- hogar_data %>% mutate(FOLIO_I = as.character(FOLIO_I))

# Merge h0327 from hogar_data into cleandata
cleandata <- left_join(cleandata, select(hogar_data, FOLIO_I, h0327), by = "FOLIO_I")

# Now safely mutate everything
cleandata <- cleandata %>%
  mutate(
    monolingual = ifelse(h0311 == 1 & h0312 == 2, 1, 0),
    tier2care = ifelse(h0309 %in% c(6, 7, 8, 11, 12), 1, 0),
    escolaridad = h0317a,
    ingreso = h0327,
    derechohabiencia = ifelse(H0310A %in% 1:9, 1, 0),
    area = ifelse(estrato == 1, 1, 0)
  )
```


## Introduction

In Mexico, healthcare disparities between Indigenous and non-Indigenous populations are a long-standing public health concern. These disparities are often attributed to systemic barriers such as geographic isolation, poverty, and discrimination in public institutions. However, one of the most persistent and underexamined drivers of healthcare inequality is *language*—specifically, the gap between healthcare institutions that operate almost exclusively in Spanish and the millions of individuals who primarily speak one of Mexico’s 68 recognized Indigenous languages. While many studies focus on the ethnic dimension of exclusion, fewer have analyzed how *variation in Spanish proficiency within Indigenous-language speakers* affects access to care.

Existing research has shown that Indigenous-language speakers face significant linguistic and cultural barriers in the Mexican healthcare system. For instance, Villarreal and Silva find that areas with higher concentrations of Indigenous-language speakers have systematically lower hospital use, even after controlling for income and rurality [@villarreal2020]. Similarly, Haimovich and Márquez Mora, studying Nahuatl-speaking communities in Puebla, found that many patients avoid hospitals not due to cost or distance, but due to an inability to communicate with medical staff and a resulting lack of trust in the system [@haimovich2021]. Aguilar-Gaxiola et al. emphasize how language exclusion intersects with social marginalization to undermine both the quantity and quality of care received by Indigenous populations [@aguilargaxiola2018].

These studies provide important evidence about the barriers faced by Indigenous communities, but they typically treat the Indigenous population as internally homogeneous. In reality, there is significant variation in Spanish fluency among Indigenous-language speakers. Some are bilingual, capable of navigating Spanish-speaking institutions, while others are *monolingual*, relying exclusively on Indigenous languages in daily life. This linguistic distinction matters. It may explain significant within-group differences in healthcare-seeking behavior that broader Indigenous vs. non-Indigenous comparisons overlook.

This project focuses specifically on this understudied divide: the difference in healthcare access between *monolingual and bilingual Indigenous-language speakers*. It uses data from the 2022 wave of the nationally representative *Encuesta Nacional de Salud y Nutrición (ENSANUT)* to examine whether individuals who speak only an Indigenous language are more likely to access *lower-tier or basic public healthcare services*, such as SSA clinics and IMSS Bienestar, compared to their bilingual peers.

Originally, the project aimed to define “informal care” narrowly—limited to traditional medicine or healers (curanderos, homeopaths, midwives). However, the data revealed that use of these providers is rare in the sample. Instead, the more revealing pattern lies in the *type and tier* of formal care accessed. While bilingual speakers access a diverse range of services—including pharmacy-affiliated clinics, private hospitals, and employer-based care—monolingual speakers are far more concentrated in basic public facilities that tend to be less administratively complex and more accessible to non-Spanish speakers.

This observation leads to a broader conceptualization of “informal” or limited-access care: not necessarily outside the formal health system, but rather at its most basic and linguistically permissive levels. These settings may be more geographically proximate and culturally familiar, but they are also often *under-resourced*, *overburdened*, and *limited in scope*. If monolingual speakers are disproportionately reliant on such facilities, it raises concerns about the equity and effectiveness of healthcare provision in Indigenous communities.

By narrowing the analysis to Indigenous-language speakers and distinguishing between monolingual and bilingual individuals, this project offers a more precise understanding of how language operates as a structural determinant of healthcare access. It also contributes to policy conversations around linguistic inclusion, suggesting that efforts to close healthcare gaps must go beyond physical access and financial protection to include *language accessibility* as a central component of equity.

In doing so, this study fills a key gap in the literature: moving beyond ethnic group comparisons to examine *within-group inequalities* based on language proficiency. It asks not just whether Indigenous people face barriers to care, but *which Indigenous people*, and why.

```{r setup, message=FALSE, warning=FALSE, echo=FALSE}
library(dplyr)
library(readr)
library(gt)

# Load datasets
cleandata <- read_csv("cleandata.csv")  # individual-level data
hogar_data <- read_csv("hogar_ensanut.csv")  # household-level income data

# Ensure FOLIO_I is a character for both
cleandata <- cleandata %>% mutate(FOLIO_I = as.character(FOLIO_I))
hogar_data <- hogar_data %>% mutate(FOLIO_I = as.character(FOLIO_I))

# Merge income data into cleandata
cleandata <- left_join(cleandata, hogar_data %>% select(FOLIO_I, h0327), by = "FOLIO_I")


cleandata <- cleandata %>%
  mutate(
    monolingual = ifelse(h0311 == 1 & h0312 == 2, 1, 0),
    tier2care = ifelse(h0309 %in% c(6, 7, 8, 11, 12), 1, 0),
    escolaridad = h0317a,
    ingreso = h0327,  # now it works because it exists!
    derechohabiencia = ifelse(H0310A %in% 1:9, 1, 0),
    area = ifelse(estrato == 1, 1, 0)
  )



```

## Theory and Hypothesis

Language is a fundamental determinant of access to public services, including healthcare. In Mexico, where the health system operates almost exclusively in Spanish, millions of Indigenous-language speakers face substantial barriers in navigating medical care. These barriers go beyond geographic or economic factors—they are linguistic and structural. For individuals who speak only an Indigenous language, the inability to communicate in Spanish can hinder their ability to understand symptoms, follow medical instructions, or even complete basic administrative procedures.

Theoretical frameworks in health access literature increasingly emphasize how language proficiency acts as a gatekeeping mechanism. Aguilar-Gaxiola et al. note that language exclusion in Mexican health services compounds socioeconomic disadvantage and contributes to underuse of institutional care [@aguilargaxiola2018]. Villarreal and Silva find that healthcare utilization rates are significantly lower in municipalities with high concentrations of Indigenous-language speakers, even after controlling for insurance coverage and rurality [@villarreal2020]. Haimovich and Márquez Mora show that linguistic mismatch between patients and providers reduces the likelihood of Indigenous-language speakers trusting and engaging with formal health services [@haimovich2021].

However, while these studies point to language as a barrier, they often treat Indigenous-language speakers as a single, homogenous group. This project builds on existing research by distinguishing between **monolingual** Indigenous-language speakers—those who do not speak Spanish—and **bilingual** speakers who do. This within-group comparison allows for a more precise examination of how language proficiency shapes healthcare behavior, particularly in settings where all respondents share Indigenous identity and cultural background.

In such a context, Spanish proficiency becomes a powerful stratifier. Bilingual speakers are more likely to navigate complex administrative systems, request services from multiple providers, and access higher-tier care. In contrast, monolingual speakers may rely on lower-tier services that are more accessible, familiar, or less administratively burdensome. These may include basic public clinics such as SSA centers and IMSS Bienestar units. Although these are part of the formal healthcare system, their administrative simplicity and geographic proximity make them more accessible to individuals with limited Spanish.

This theoretical perspective views language not only as a communication tool but as a mechanism of **institutional accessibility**. In the absence of Spanish proficiency, monolingual Indigenous speakers are structurally disadvantaged in their ability to access diverse, high-quality care. Their limited provider options may not reflect personal choice, but rather constrained access rooted in linguistic exclusion.

> **Hypothesis**: Among Indigenous-language speakers in Mexico, individuals who do not speak Spanish (monolinguals) are more likely to rely on lower-tier or basic public healthcare services than bilingual speakers, due to the linguistic inaccessibility of more complex care institutions.

This hypothesis is both falsifiable and policy-relevant. If monolingual speakers are found to use a narrower and less institutionally diverse range of providers, it would suggest that improving language accessibility should be a central focus in addressing Indigenous health disparities. Policies such as expanded translation services, Indigenous-language outreach, and training of bilingual medical staff would be critical steps toward equitable care.


## Research Design

This analysis draws on data from the 2022 wave of the *Encuesta Nacional de Salud y Nutrición Continua* (ENSANUT), administered by Mexico’s *Instituto Nacional de Salud Pública (INSP)*. ENSANUT is a nationally representative cross-sectional survey that uses a probabilistic, stratified sampling design to produce reliable health and demographic estimates at the national, regional, and urban/rural levels. The dataset includes rich information on healthcare usage, language use, education, household income, and access to insurance, making it suitable for examining how structural factors influence healthcare-seeking behavior.

The **unit of analysis** is the individual respondent. For this study, the sample is restricted to adults aged 18 and older who report speaking an Indigenous language, using item **H0311** (*¿Habla alguna lengua indígena?*). While ENSANUT does not directly ask about ethnic identity, both INSP and Mexico’s national statistics agency (INEGI) routinely use Indigenous language use as a proxy for Indigenous identity in population-level studies [@inegi2019]. This approach is standard in health equity research and allows for robust analysis of intra-group differences within Indigenous populations.

Within this sample, respondents are divided into two groups using item **H0312** (*¿Habla también español?*):
- **Monolinguals**: Speak an Indigenous language but not Spanish (H0311 = yes, H0312 = no).
- **Bilinguals**: Speak both an Indigenous language and Spanish (H0311 = yes, H0312 = yes).

A binary variable `monolingual` is created to distinguish these groups, coded as 1 for monolinguals and 0 for bilinguals. This distinction allows us to test whether the absence of Spanish proficiency, independent of shared Indigenous identity, influences patterns of healthcare access.

The **dependent variable** is the type of healthcare institution the respondent usually visits when facing a health need. Item **H0309** asks: *“Cuando tiene una necesidad de salud, ¿en dónde se atiende usualmente?”* The responses are grouped into a **three-tier typology** based on institutional complexity and accessibility:
- **Tier 1 (High-Complexity Institutional Care)**: IMSS, ISSSTE, PEMEX, Defense, private hospitals.
- **Tier 2 (Basic Public Care)**: SSA clinics, IMSS Bienestar, and DIF.
- **Tier 3 (Non-Institutional/Alternative Care)**: pharmacy consults, home-based care, traditional healers, homeopaths.

For the purposes of hypothesis testing, a binary variable `tier2_care` is constructed: 1 if the respondent uses Tier 2 care, 0 if they use any other tier. Tier 2 institutions are the most common formal care sites in underserved areas and represent the **lowest-barrier access point** in the public system. They do not require employment-based affiliation (as with IMSS) and are generally easier to navigate for individuals with limited literacy or language proficiency. Focusing on Tier 2 allows us to test whether monolingual speakers are **funneled into this tier** as a result of structural and linguistic barriers that restrict access to Tier 1 care, without resorting to informal or alternative care sources (Tier 3).

To reduce omitted variable bias, the model includes the following **control variables**:
- **Education level (H0317A)**: Ordinal variable capturing highest grade completed (no schooling to university).
- **Income bracket**: Derived from ENSANUT’s household income sections (I and VI), capturing relative socioeconomic status.
- **Insurance status (H0310A)**: Whether the respondent has formal healthcare coverage through IMSS, ISSSTE, or other systems.
- **Urban/rural residence (estrato)**: Binary variable indicating household location type.

# Summary Statistics

```{r summary, message=FALSE, warning=FALSE, echo=FALSE}
library(dplyr)
library(gt)
library(readr)

# Load datasets
cleandata <- read_csv("cleandata.csv")
hogar_data <- read_csv("hogar_ensanut.csv")

# Merge income variable (h0327) into cleandata using FOLIO_I
cleandata <- cleandata %>%
  mutate(FOLIO_I = as.character(FOLIO_I))
hogar_data <- hogar_data %>%
  mutate(FOLIO_I = as.character(FOLIO_I))

cleandata <- left_join(cleandata, hogar_data %>% select(FOLIO_I, h0327), by = "FOLIO_I")

# Define analysis variables
cleandata <- cleandata %>%
  mutate(
    monolingual = ifelse(h0311 == 1 & h0312 == 2, 1, 0),
    tier2care = ifelse(h0309 %in% c(6, 7, 8, 11, 12), 1, 0),
    escolaridad = h0317a,
    ingreso = h0327,
    derechohabiencia = ifelse(H0310A %in% 1:9, 1, 0),
    area = ifelse(estrato == 1, 1, 0)
  )
```

```{r summary-stats-final, message=FALSE, warning=FALSE, echo=FALSE}
library(dplyr)
library(gt)
library(tidyr)


# Ensure tier2care is numeric (0/1)
cleandata$tier2care <- as.numeric(cleandata$tier2care)

# Select only the variables of interest
summary_data <- cleandata %>%
  select(monolingual, tier2care, escolaridad, ingreso, derechohabiencia, area) %>%
  summarise(across(
    everything(),
    list(
      Mean = ~mean(., na.rm = TRUE),
      SD = ~sd(., na.rm = TRUE),
      Min = ~min(., na.rm = TRUE),
      Max = ~max(., na.rm = TRUE)
    ),
    .names = "{.col}_{.fn}"
  )) %>%
  pivot_longer(cols = everything(),
               names_to = c("Variable", "Statistic"),
               names_sep = "_") %>%
  pivot_wider(names_from = Statistic, values_from = value) %>%
  gt() %>%
  tab_header(title = "Table 1. Summary Statistics of Key Variables") %>%
  fmt_number(columns = c(Mean, SD, Min, Max), decimals = 2)

summary_data
```






These controls account for structural drivers of healthcare access—such as affordability, geographic proximity, and institutional affiliation—that could confound the relationship between language ability and healthcare-seeking behavior.

The main model is a logistic regression estimating the probability of Tier 2 care usage:

```{r model, message=FALSE, warning=FALSE, echo=FALSE}
#| echo: false
#| warning: false
#| message: false
glm(tier2care ~ monolingual + escolaridad + ingreso + derechohabiencia + area,
    family = binomial, data = cleandata)
```

### Empirical Extension

To test the robustness of the relationship, we conduct an empirical extension by restricting the sample to respondents with **mid-level income** and **completed primary school**. This subgroup analysis helps address the possibility that language ability is merely proxying for deeper structural exclusion. If monolingual speakers are still more likely to use informal care within a relatively homogeneous SES subgroup, it would strengthen the argument that **language accessibility alone**—not poverty or education—is driving the observed effect.

This approach allows us to more confidently interpret the coefficient on `only_indig_lang` as a function of linguistic exclusion, rather than as an artifact of correlated disadvantage.


```{r oaxaca-extension, message=FALSE, warning=FALSE, echo=FALSE}
#| echo: false
#| warning: false
#| message: false
oaxaca_model <- glm(tier2care ~ monolingual + escolaridad + ingreso + derechohabiencia + area,
                    data = filter(cleandata, desc_ent1 == "20 OAXACA"),
                    family = binomial)

summary(oaxaca_model)
```

```{r cdmx-extension, message=FALSE, warning=FALSE, echo=FALSE}

#| echo: false
#| warning: false
#| message: false
# Logistic regression limited to Ciudad de México (state code "09 CIUDAD DE MÉXICO")
cdmx_model <- glm(tier2care ~ monolingual + escolaridad + ingreso + derechohabiencia + area,
                  data = filter(cleandata, desc_ent1 == "09 CIUDAD DE MÉXICO"),
                  family = binomial)

summary(cdmx_model)
```


### Findings


```{r regression-table, message=FALSE, warning=FALSE, echo=FALSE}
# Load necessary package
library(modelsummary)

# Fit the original model
og_model <- glm(tier2care ~ monolingual + escolaridad + ingreso + derechohabiencia + area,
                family = binomial, data = cleandata)

# Display the pretty regression summary table
modelsummary(og_model,
             title = "Table 1. Logistic Regression: Predicting Use of Tier 2 Care",
             estimate = "{estimate}{stars}",
             statistic = "({std.error})",
             gof_omit = "AIC|BIC|Log.Lik|Deviance|RMSE")
```





```{r top-care-barplot, message=FALSE, warning=FALSE, echo=FALSE}
library(dplyr)
library(ggplot2)
library(forcats)

# Map numeric codes from h0309 to readable care types
care_labels <- c(
  `1` = "IMSS", `2` = "ISSSTE Estatal", `3` = "PEMEX", `4` = "Defensa",
  `5` = "Marina", `6` = "SSA Clinic", `7` = "IMSS Bienestar",
  `8` = "DIF", `9` = "Red Cross", `10` = "National Institute",
  `11` = "Other NGO", `12` = "Pharmacy Clinic", `13` = "Private Hospital",
  `14` = "Hospital ER", `15` = "Private Tower", `16` = "Private Home Visit",
  `17` = "Own Home Visit", `18` = "Telemedicine", `19` = "Occupational Doctor",
  `20` = "Curandero/Naturista", `21` = "Homeopath/Midwife", `22` = "Other",
  `23` = "No response"
)

# Create label column for chart
cleandata <- cleandata %>%
  mutate(
    care_label = recode(as.character(h0309), !!!care_labels),
    Language_Group = ifelse(monolingual == 1, "Monolingual", "Bilingual")
  )

# Count by care type and language group
care_plot_data <- cleandata %>%
  filter(!is.na(care_label)) %>%
  count(Language_Group, care_label) %>%
  group_by(Language_Group) %>%
  mutate(prop = n / sum(n)) %>%
  ungroup()

# Optional: filter for most common types
top_care <- care_plot_data %>%
  group_by(care_label) %>%
  summarise(total = sum(n)) %>%
  arrange(desc(total)) %>%
  slice_max(total, n = 12) %>%
  pull(care_label)

# Plot
ggplot(
  filter(care_plot_data, care_label %in% top_care),
  aes(x = fct_reorder(care_label, -prop), y = prop, fill = Language_Group)
) +
  geom_col(position = "dodge") +
  coord_flip() +
  labs(
    title = "Top Healthcare Sources by Language Group (ENSANUT 2022)",
    x = "Usual Place of Care",
    y = "Proportion of Respondents",
    fill = "Language Group"
  ) +
  theme_minimal(base_size = 13)
```



```{r regression-tablesf, message=FALSE, warning=FALSE, echo=FALSE}
library(modelsummary)

# Run the three models
model_national <- glm(tier2care ~ monolingual + escolaridad + ingreso + derechohabiencia + area,
                      data = cleandata, family = binomial)

model_oaxaca <- glm(tier2care ~ monolingual + escolaridad + ingreso + derechohabiencia + area,
                    data = filter(cleandata, desc_ent1 == "20 OAXACA"), family = binomial)

# Combine three major states: Jalisco, CDMX, Nuevo León
target_states <- c("14 JALISCO", "09 CIUDAD DE MÉXICO", "19 NUEVO LEÓN")

# Filter data
combined_states_data <- cleandata %>%
  filter(desc_ent1 %in% target_states)

# Run the regression model
combined_model <- glm(tier2care ~ monolingual + escolaridad + ingreso +
                        derechohabiencia + area,
                      family = binomial,
                      data = combined_states_data)


# Display models in a comparative table
modelsummary(
  list("National" = model_national,
       "Oaxaca" = model_oaxaca,
       "Jalisco, CDMX & Nuevo León" = combined_model),
  fmt = 3,
  stars = TRUE,
  title = "Table: Effect of Monolingualism on Use of Tier 2 Healthcare"
)
```

------------------------------------------------------------------------
